Inferring generation-interval distributions from contact tracing data: \\ \emph{in prep, PRSB}

### 2020-05-23 doc/response.tex

\revtext{
I also recommend, for the authors to consider, breaking this paper
into two papers: one on mathematical theory and models to better
clarify concepts which I believe are still ambiguous in currently published literature; and another one on statistical models and methods
to link data to models.
}

\reviewer{Reviewer \#2}

\revtext{
This manuscript extends previous work to examine how observed generation intervals (estimated through contract tracing) are shaped by both temporal and spatial effects during the course of an epidemic and leverages this to construct two methods that correct for the time-censoring bias. These methods are then applied to simulated data with complete reporting and shown to accurately infer the reproductive number from the exponential growth rate.
}

\revtext{
The manuscript is well-written and presents novel methods to correct observed generation intervals to obtain estimates of R0. When reading sections 3 and 4, I was wondering if there was a conceptual diagram that would help the reader visualize the different spatio-temporal processes that impact the observed generation-interval distributions (summarized on lines 224-226).}

\revtext{Finally, while the methods are tested on simulated data with complete reporting, it's unclear how they would perform with incomplete reporting, and ultimately, with empirical data. Of course, the latter is beyond the scope of the current manuscript but it would be useful to include a straightforward treatment of simulated data with incomplete reporting to support the expectations discussed on lines 265-267.}

We have added a new figure to the supplementary figures to account for incomplete reporting.

### 2020-05-23 doc/network.tex
tweaks

\begin{figure}[!pbth]
\includegraphics[width=\textwidth]{../fig/individual_and_population.pdf}
\caption{\textbf{Comparison of individual- and population-level kernels.}
(Left) an individual-level kernel of an infected individual with latent period of 11.4 days followed by infectious period of 5 days;
this represents an individual realization of a random process.
(Right) a population-level kernel of infected individuals with latent and infectious periods exponentially distributed with means of 11.4 and 5 days, respectively;
this represents a population average of a random process.
Shaded areas under the curves are equal to individual- and population-level reproductive numbers, both of which are set to 2 in this example.
Parameters are chosen to reflect the West African Ebola outbreak \citep{who2014ebola}.}
\label{fig:indpop}
\end{figure}

@@

\begin{figure}[!pbth]
\includegraphics[width=\textwidth]{../fig/temporal_effect.pdf}
\caption{\textbf{Temporal variation in the mean observed generation interval.}
A deterministic Susceptible-Exposed-Infectious-Recovered (SEIR) model was simulated using Ebola-like parameters \citep{who2014ebola}: mean latent period $1/\sigma = 11.4 \textrm{ days}$, mean infectious period $1/\gamma = 5 \textrm{ days}$, and the basic reproductive number $\RR_0 = 2$. 
The backward and the aggregated mean generation interval were calculated over the course of an epidemic.
The dotted horizontal line represents the mean intrinsic generation interval.
}
\label{fig:censor}
\end{figure}

@@

\begin{figure}[!pbth]
\includegraphics[width=\textwidth]{../fig/local_effect.pdf}
\caption{
\textbf{Spatial effects on realized generation intervals.}
Theoretical distributions and means are shown in color (and are the same in each panel, for reference). Simulated distributions and means are shown in black.
(Left) the intrinsic generation-interval distribution corresponds to all contacts by a focal individual, regardless of whether the individual contacted is susceptible.
(Middle) the egocentric generation-interval distribution corresponds to the distribution of all infectious contacts by the focal individual with susceptible individuals, in the case where the focal individual is the only possible infector (simulated on a star network).
(Right) realized generation-interval distributions are shorter than egocentric distributions in general, because contacts can be wasted when susceptibles become infected through other routes (simulated on a homogeneous network).
All figures were generated using 5000 stochastic simulations on a network with 5 nodes (1 infector and 4 susceptibles) with Ebola-like parameters \citep{who2014ebola}:
mean latent period $1/\sigma = 11.4 \textrm{ days}$ and mean infectious period $1/\gamma = 5 \textrm{ days}$. 
Per-pair contact rate $\lambda = 0.25 \textrm{ days}^{-1}$ is chosen to be sufficiently high so that the differences between generation-interval distributions are clear.
Each simulation is run until all individuals are either susceptible or have recovered.
}
\label{fig:local}
\end{figure}

@@

When realized generation intervals are aggregated over the course of an epidemic, there will be four effects present in the sample: (1) right-truncation effect, (2) egocentric depletion effect, (3) local depletion effect and (4) global depletion effect.
We can correct explicitly for the egocentric effect and, in the case of exponential growth, the right-truncation effect;
these effects shorten the mean realized generation intervals, which in turn will reduce the estimate of the reproductive number \cite{wallinga2007generation, park2019practical}.
While the other two effects are difficult to measure, we can make qualitative predictions about their effects on the realized generation intervals and reproductive numbers: 
both local and global depletion effects also reduce number of infections that occur and shorten generation intervals.
If we can correct for the temporal bias early in an outbreak, during the exponential growth phase, we should be able to infer the initial forward generation-interval distribution, which incorporates egocentric and local spatial effects but not the global effects, from the aggregated distribution.

\begin{figure}[!pbth]
\includegraphics[width=\textwidth]{../fig/cmp_reproductive.pdf}
\caption{\textbf{Comparison of estimates of reproductive number based on various methods.}
Using the observed generation-interval distributions (based on the first 1000 generation intervals) without correcting for right-truncation severely underestimates the reproductive number.
Similarly, using the intrinsic generation-interval distribution overestimates the reproductive number because it fails to account for local spatial effects; the egocentric distribution corrects for this only partially.
Both population-level and individual-level methods provide estimates of reproductive number that are consistent with the empirical estimates, which we define as the average number of secondary cases generated by the first 100 infected individuals, but the individual-level method is more precise.
Boxplots are generated using 100 stochastic simulations of the SEIR model on an empirical network using Ebola-like parameters \citep{who2014ebola}: mean latent period $1/\sigma = 11.4 \textrm{ days}$ and mean infectious period $1/\gamma = 5 \textrm{ days}$. 
Per-pair contact rate $\lambda = 0.08 \textrm{ days}^{-1}$ is chosen to be sufficiently high such that differences are clear.
}
\label{fig:cmp}
\end{figure}

>>>>

\begin{figure}[!ht]
\includegraphics[width=\textwidth]{../fig/individual_and_population.pdf}
\caption{\textbf{Comparison of individual- and population-level kernels.}
(Left) an individual-level kernel of an infected individual with latent period of 11.4 days followed by infectious period of 5 days;
this represents an individual realization of a random process.
(Right) a population-level kernel of infected individuals with latent and infectious periods exponentially distributed with means of 11.4 and 5 days, respectively;
this represents a population average of a random process.
Shaded areas under the curves are equal to individual- and population-level reproductive numbers, both of which are set to 2 in this example.
Parameters are chosen to reflect the West African Ebola outbreak \citep{who2014ebola}.}
\label{fig:indpop}
\end{figure}

@@

\begin{figure}[!ht]
\includegraphics[width=\textwidth]{../fig/temporal_effect.pdf}
\caption{\textbf{Temporal variation in the mean backward and aggregated generation interval.}
A deterministic Susceptible-Exposed-Infectious-Recovered (SEIR) model is simulated using Ebola-like parameters \citep{who2014ebola}: mean latent period $1/\sigma = 11.4 \textrm{ days}$, mean infectious period $1/\gamma = 5 \textrm{ days}$, and the basic reproductive number $\RR_0 = 2$. 
The backward and the aggregated mean generation interval are calculated over the course of an epidemic.
The dotted horizontal line represents the mean intrinsic generation interval.
}
\label{fig:censor}
\end{figure}

@@

\begin{figure}[!ht]
\includegraphics[width=\textwidth]{../fig/local_effect.pdf}
\caption{
\textbf{Spatial effects on realized generation intervals.}
Theoretical distributions and means are shown in color (and are the same in each panel, for reference). Simulated distributions and means are shown in black.
(Left) the intrinsic generation-interval distribution corresponds to all contacts by a focal individual, regardless of whether the individual contacted is susceptible.
(Middle) the egocentric generation-interval distribution corresponds to the distribution of all infectious contacts by the focal individual with susceptible individuals, in the case where the focal individual is the only possible infector (simulated on a star network).
(Right) realized generation-interval distributions are shorter than egocentric distributions in general, because contacts can be wasted when susceptibles become infected through other routes (simulated on a homogeneous network).
All figures were generated using 5000 stochastic simulations on a network with 5 nodes (1 infector and 4 susceptibles) with Ebola-like parameters \citep{who2014ebola}:
mean latent period $1/\sigma = 11.4 \textrm{ days}$ and mean infectious period $1/\gamma = 5 \textrm{ days}$. 
Per-pair contact rate $\lambda = 0.25 \textrm{ days}^{-1}$ is chosen to be sufficiently high so that the differences between generation-interval distributions are clear.
Each simulation is run until all individuals are either susceptible or have recovered.
}
\label{fig:local}
\end{figure}

@@

\begin{figure}[!ht]
\includegraphics[width=\textwidth]{../fig/diagram.pdf}
\caption{
\textbf{Spatiotemporal effects on generation intervals.}
}
\label{fig:diagram}
\end{figure}

When realized generation intervals are aggregated over the course of an epidemic, there will be four effects present in the sample (\fref{diagram}): (1) right-truncation effect, (2) egocentric depletion effect, (3) local depletion effect and (4) global depletion effect.
We can correct explicitly for the egocentric effect and, in the case of exponential growth, the right-truncation effect;
these effects shorten the mean realized generation intervals, which in turn will reduce the estimate of the reproductive number \cite{wallinga2007generation, park2019practical}.
While the other two effects are difficult to measure, we can make qualitative predictions about their effects on the realized generation intervals and reproductive numbers: 
both local and global depletion effects also reduce number of infections that occur and shorten generation intervals.
If we can correct for the temporal bias early in an outbreak, during the exponential growth phase, we should be able to infer the initial forward generation-interval distribution, which incorporates egocentric and local spatial effects but not the global effects, from the aggregated distribution.

\begin{figure}[!ht]
\includegraphics[width=\textwidth]{../fig/cmp_reproductive.pdf}
\caption{\textbf{Comparison of estimates of reproductive number based on various methods.}
Using the observed generation-interval distributions (based on the first 1000 generation intervals) without correcting for right-truncation severely underestimates the reproductive number.
Similarly, using the intrinsic generation-interval distribution overestimates the reproductive number because it fails to account for local spatial effects; the egocentric distribution corrects for this only partially.
Both population-level and individual-level methods provide estimates of reproductive number that are consistent with the empirical estimates, which we define as the average number of secondary cases generated by the first 100 infected individuals, but the individual-level method is more precise.
Boxplots are generated using 100 stochastic simulations of the SEIR model on an empirical network using Ebola-like parameters \citep{who2014ebola}: mean latent period $1/\sigma = 11.4 \textrm{ days}$ and mean infectious period $1/\gamma = 5 \textrm{ days}$. 
Per-pair contact rate $\lambda = 0.08 \textrm{ days}^{-1}$ is chosen to be sufficiently high such that differences are clear.
}
\label{fig:cmp}
\end{figure}
